---
title: "Project proposal"
author: "Rat Bastards"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

## 1. Introduction

Our data are a collection of phenological observations on vegetation from Hurricane Island. Phenology is the study of the changes in development that biological organisms progress through with the seasons. This includes phenomenona such as leafing out, budding, flowering, fruiting, and leaves turning and falling. These phenomenona usually occur sequentially depending on a variety of factors such as precipitation and amount of days above a minimum temperature. Each plant has different needs for development, and these needs are important to be met so that their timings align with other phenomena such as pollinator timing and seeds falling at the proper time for regeneration. This dataset was gathered in the 2023 spring-fall field season to track the phenophases of a broad range of plant species. The data were collected by one of our team members who visited individual plants each week throughout the season. Photos were taken of each plant on a weekly basis to identify phenophase development. The phenophases tracked were based on the categories described by the National Phenology Network. The primary goal of this research is to create and visualize a baseline phenological record of the vegetation on Hurricane Island which can be used in future analysis. Additionally, we plan to analyze how the phenophases develop in response to varying climatic factors throughout the season. We also intend to visualize comparisons between differing species categories such as fruit-bearing/non-fruit-bearing, native/introduced, etc to see how their development varies. The cases are each plant species and its disposition on the calendar day on which it was observed. The variables are date of data collection, species information, and phenological observations made. Individual species have differing phenological observations made depending on varying morphology. (Eg. A plant that fruits versus one that does not).


## 2. Data

```{r hurricane-phenology-data, warning=FALSE}
library(readr)

hurricane_raw_data <- read_csv("../data/hurricane_plants.csv")

# remove empty rows, where life_form is NA 
hurricane_raw_data <- hurricane_raw_data %>% 
  filter(!is.na(life_form))
# another option for the same function: drop_na(life_form)

# reorder variables to logical order
hurricane_plants <- relocate(hurricane_raw_data, 
       date, 
       life_form, 
       species, 
       initial_emergence,
       breaking_leaf_buds_count,
       leaf_presence,
       percent_unfolded_leaves,
       unfolded_leaves_count,
       percent_full_size_leaf,
       percent_leaves_colorful,
       fallen_leaf_presence,
       breaking_needle_bud_count,
       young_needle_count,
       percent_stalk_growth,
       percent_fiddlehead_unrolled,
       buds_and_flowers_count,
       percent_open_flowers,
       pollen_amount,
       pollen_cone_count,
       percent_open_pollen_cones,
       fruit_count,
       unripe_seed_cone_count, 
       percent_ripe_fruits,
       ripe_seed_cone_count,
       dropped_fruit_count,
       notes
       )

# write dataframe to CSV file
# write_csv(hurricane_plants, file = "hurricane_plants_clean.csv")

glimpse(hurricane_plants)
```

```{r raster-package}
# install.packages("raster")
library(raster)
library(tidyverse)
# note we may need to change to the geodata package
```


```{r neracoos_buoy_data}
# load weather data from neracoos
weather_data <- read_csv("../data/neracoos_buoy_data.csv")
```

## 3. Ethics review
Data was collected by hand by one of our group members in an noninvasive way.

## 4. Data analysis plan

We will visualize date, breaking leaves/buds/needles, flowering, and fruiting phenophases of comparable species. We will also visualize these phenophases based on records of climatic variables from this field season. The climatic variables will need to be sourced from the NERACOOS database. We plan on utilizing violin plots, lollipop plots, ridge plots, as well as potential animations and maps.
```{r summary-stats-distinct}
# how many plants (total)
distinct(hurricane_plants, species) #there are 24 individual species
```

There are 24 individual species. This is a large number to see individual values on a plot. Next, we will see how many individuals are classified in each life form group:

```{r distinct-life-forms}
# how many life forms?
distinct(hurricane_plants, life_form) #there are 4 life form types 
# types are tree,	shrub, vine, and herb

# how many trees?
hurricane_plants %>%
  filter(life_form == "tree") %>%
  distinct(species) # there are 3 trees

# how many shrubs?
hurricane_plants %>%
  filter(life_form == "shrub") %>%
  distinct(species) # there are 8 shrubs

# how many vines?
hurricane_plants %>%
  filter(life_form == "vine") %>%
  distinct(species) # there is 1 vine

# how many herbs?
hurricane_plants %>%
  filter(life_form == "herb") %>%
  distinct(species) # there are 11 herbs
```


```{r summary-stats-max-min-breaking-leaf-and-needle}
hurricane_plants %>%
  group_by(species) %>%
   # filter(!is.na(breaking_leaf_buds_count),
   #        !is.na(breaking_needle_bud_count)) %>%
  summarise(max_breaking_leaf = max(breaking_leaf_buds_count, na.rm = T),
            max_breaking_needle = max(breaking_needle_bud_count, na.rm = T)) #not returning breaking needle count for the red spruce. Something wrong with dataset?
```

```{r summary-stats-initial-emergence}
hurricane_plants %>%
  group_by(life_form) %>%
  filter(initial_emergence == T) %>%
  summarise(initial_emergence_date_min = min(date),
            initial_emergence_date_max = max(date))

hurricane_plants %>%
  filter(life_form == "tree",
         breaking_leaf_buds_count > 0) %>%
  summarise(tree_initial_emergence_date_min = min(date),
            tree_initial_emergence_date_max = max(date))
# more detailed initial growth data located in analysis file
```
```{r summary-stats-leaf_emergence}
# calculate maximum breaking leaf buds
hurricane_plants %>%
  group_by(species) %>%
  filter(!is.na(breaking_leaf_buds_count)) %>%  #remove NA values
  summarise(max_leaves_breaking = max(breaking_leaf_buds_count)) 
# why are there only 9 observations?

hurricane_plants %>%
  group_by(life_form) %>%
  filter(!is.na(unfolded_leaves_count)) %>%
  summarise(average_unfolded_leaves = mean(unfolded_leaves_count)) 
```


```{r flowers_and_buds, warning=FALSE}
# graph the amount of buds or flowers of each plant? facet by life form?
hurricane_plants %>%
  mutate(date = mdy(date)) %>%
ggplot(mapping = 
         aes(x = date, y = buds_and_flowers_count, color = species)) +
  geom_point() +
  # geom_density() +
  facet_wrap(~ life_form) +
  labs (y = "flowers and buds",
        title = "amount of buds and flowers by life form over time")
```


```{r total_fruits, warning=FALSE}
# graph total fruits over time
hurricane_plants %>%
  mutate(date = mdy(date)) %>%
  group_by(date) %>%
  summarise(total_fruits = sum(fruit_count, na.rm = T)) %>%
  ggplot(aes(
    x = date,
    y = total_fruits)) +
 geom_point() +
  labs(x = "date",
       y = "total fruits",
       title = "cumulative fruits with seasonal progression") +
  geom_line()

```

The dips in late June and early July are interesting. We suspect that perhaps not all species were accounted for on these dates but will need to look deeper into our data.
